---
title: "Human and financial cost of weather events across USA"
author: "Tiago Fonseca"
date: "21-09-2014"
output: html_document
---

```{r global_options, include=FALSE}
	opts_chunk$set(fig.width=12, fig.height=8, cache=FALSE, warning=FALSE)
```

## Abstract

In this documents we study the fatalities, injuries and financial damages caused by weather events.
We focus ourselves in two different set of questions:

1. Are amount of fatalities, injuries and the financial cost correlated?

2. Which are the events that are more expensive in terms of fatalities, injuries and financial damages?

In order to achieve our goal, we load the file from the indicated url, and we clean up the data.
In the data there is two sorts of financial damages and we add one to the other.
Then, we compute the mean and the sum for the three concerned quantities.

There is a lot of mistakes on the strings indicating the type of event, then we filter all the small things (small amount of injuries and fatalities and financial damages).
So keep in mind that some data is removed from our anlysis. 
The right thing to do would be to read all the types and merge them together when necessary, but this would be a very time consuming process.

Finally, we plot the correlation between the average and the ranking according to the mean and the sum.

## Data Processing 

### Loading data

The data was obtained directly from the indicated url, and it was stored in the memory.

```{r message = FALSE}
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
                "Storm_data.csv.bz2", method = "curl")
  data <- read.table("Storm_data.csv.bz2", sep = ",", head = TRUE, na.strings = "")
```

### Cleaning data

In order to clean the data, we remove the NA's. 
The amount of fatalities, injuries and damages if not present is replaced by 0.

```{r}
  data$CROPDMG <- ifelse(is.na(data$CROPDMG), 0, data$CROPDMG)
  data$PROPDMG <- ifelse(is.na(data$PROPDMG), 0, data$PROPDMG)
  data$INJURIES <- ifelse(is.na(data$INJURIES), 0, data$INJURIES)
  data$FATALITIES <- ifelse(is.na(data$FATALITIES), 0, data$FATALITIES)
```

Let us tranform the exponents in readable numbers, first the crop damages:
```{r}
  summary(data$CROPDMGEXP)
  data$CROPDMGEXP <- ifelse(is.na(data$CROPDMGEXP), 1,
                ifelse(data$CROPDMGEXP == "B", 1E9,
                ifelse(data$CROPDMGEXP == "M", 1E6,
                ifelse(data$CROPDMGEXP == "m", 1E6,
                ifelse(data$CROPDMGEXP == "K", 1E3,
                ifelse(data$CROPDMGEXP == "k", 1E3,
                ifelse(data$CROPDMGEXP == "2", 1E2, 1)))))))
```

And the property damages:
```{r}
  summary(data$PROPDMGEXP)
  data$PROPDMGEXP <- ifelse(is.na(data$PROPDMGEXP), 1,
                ifelse(data$PROPDMGEXP == "B", 1E9,
                ifelse(data$PROPDMGEXP == "M", 1E6,
                ifelse(data$PROPDMGEXP == "m", 1E6,
                ifelse(data$PROPDMGEXP == "K", 1E3,
                ifelse(data$PROPDMGEXP == "H", 1E2,
                ifelse(data$PROPDMGEXP == "h", 1E2,
                ifelse(data$PROPDMGEXP == "8", 1E8,
                ifelse(data$PROPDMGEXP == "7", 1E7,
                ifelse(data$PROPDMGEXP == "6", 1E6,
                ifelse(data$PROPDMGEXP == "5", 1E5,
                ifelse(data$PROPDMGEXP == "4", 1E4,
                ifelse(data$PROPDMGEXP == "3", 1E3,
                ifelse(data$PROPDMGEXP == "2", 1E2,
                ifelse(data$PROPDMGEXP == "1", 1E1, 1)))))))))))))))
```


The multiplicative factor for damage if not present is replaced by 1.

In order to simplify the computations all event types were transformed to capital letters. 
And gather them in 40 categories.

```{r}
  data$EVTYPE <- toupper(data$EVTYPE)
  data$EVTYPE[grep("HAIL", data$EVTYPE)] <- "HAIL"
  data$EVTYPE[grep("TORNADO", data$EVTYPE)] <- "TORNADO"
  data$EVTYPE[grep("SPOUT", data$EVTYPE)] <- "TORNADO"
  data$EVTYPE[grep("TORNDAO", data$EVTYPE)] <- "TORNADO"
  data$EVTYPE[grep("HURRICANE", data$EVTYPE)] <- "HURRICANE"
  data$EVTYPE[grep("FLOOD", data$EVTYPE)] <- "FLOOD"
  data$EVTYPE[grep("THUNDER", data$EVTYPE)] <- "THUNDERSTORM"
  data$EVTYPE[grep("GUSTNADO", data$EVTYPE)] <- "THUNDERSTORM"
  data$EVTYPE[grep("TSTM", data$EVTYPE)] <- "THUNDERSTORM"
  data$EVTYPE[grep("TIDE", data$EVTYPE)] <- "TIDES"
  data$EVTYPE[grep("WIND", data$EVTYPE)] <- "WIND"
  data$EVTYPE[grep("WND", data$EVTYPE)] <- "WIND"
  data$EVTYPE[grep("FIRE", data$EVTYPE)] <- "FIRE" 
  data$EVTYPE[grep("RAIN", data$EVTYPE)] <- "RAIN"
  data$EVTYPE[grep("SNOW", data$EVTYPE)] <- "SNOW"
  data$EVTYPE[grep("SLEET", data$EVTYPE)] <- "SNOW"
  data$EVTYPE[grep("FOG", data$EVTYPE)] <- "FOG" 
  data$EVTYPE[grep("ICE", data$EVTYPE)] <- "ICE"
  data$EVTYPE[grep("ICY", data$EVTYPE)] <- "ICE"
  data$EVTYPE[grep("GLAZE", data$EVTYPE)] <- "ICE"
  data$EVTYPE[grep("COLD", data$EVTYPE)] <- "COLD"
  data$EVTYPE[grep("FREEZ", data$EVTYPE)] <- "COLD"
  data$EVTYPE[grep("FROST", data$EVTYPE)] <- "COLD"
  data$EVTYPE[grep("LOW TEMP", data$EVTYPE)] <- "COLD"
  data$EVTYPE[grep("HEAT", data$EVTYPE)] <- "HEAT"
  data$EVTYPE[grep("HOT", data$EVTYPE)] <- "HEAT"
  data$EVTYPE[grep("WARM", data$EVTYPE)] <- "HEAT"
  data$EVTYPE[grep("HIGH TEMP", data$EVTYPE)] <- "HEAT"
  data$EVTYPE[grep("DRY", data$EVTYPE)] <- "DROUGHT"
  data$EVTYPE[grep("DROUGHT", data$EVTYPE)] <- "DROUGHT"
  data$EVTYPE[grep("DRIEST", data$EVTYPE)] <- "DROUGHT"
  data$EVTYPE[grep("DUST", data$EVTYPE)] <- "DUST" 
  data$EVTYPE[grep("STORM", data$EVTYPE)] <- "STORM"
  data$EVTYPE[grep("SHOWER", data$EVTYPE)] <- "RAIN"
  data$EVTYPE[grep("PRECI",data$EVTYPE)] <- "RAIN"
  data$EVTYPE[grep("WINTER",data$EVTYPE)] <- "RAIN"
  data$EVTYPE[grep("WINTRY",data$EVTYPE)] <- "RAIN"
  data$EVTYPE[grep("LIGHT", data$EVTYPE)] <- "LIGHTNING"
  data$EVTYPE[grep("WET", data$EVTYPE)] <- "WETNESS" 
  data$EVTYPE[grep("FLOOODING", data$EVTYPE)] <- "FLOOD" 
  data$EVTYPE[grep("FLD", data$EVTYPE)] <- "FLOOD" 
  data$EVTYPE[grep("STREAM", data$EVTYPE)] <- "FLOOD" 
  data$EVTYPE[grep("SURF", data$EVTYPE)] <- "SURF ADVISORY"
  data$EVTYPE[grep("AVA", data$EVTYPE)] <- "AVALANCHE"
  data$EVTYPE[grep("VOLCANIC", data$EVTYPE)] <- "VOLCANIC EVENT"
  data$EVTYPE[grep("VOG", data$EVTYPE)] <- "VOLCANIC EVENT"
  data$EVTYPE[grep("SLIDE", data$EVTYPE)] <- "LANDSLIDE"
  data$EVTYPE[grep("SLUMP", data$EVTYPE)] <- "LANDSLIDE"
  data$EVTYPE[grep("EROSION", data$EVTYPE)] <- "COASTAL EROSION"
  data$EVTYPE[grep("EROSIN", data$EVTYPE)] <- "COASTAL EROSION"
  data$EVTYPE[grep("COASTAL", data$EVTYPE)] <- "COASTAL EROSION"
  data$EVTYPE[grep("RIP", data$EVTYPE)] <- "RIP CURRENT"
  data$EVTYPE[grep("BLIZZARD", data$EVTYPE)] <- "BLIZZARD"
  data$EVTYPE[grep("CLOUD", data$EVTYPE)] <- "WALL/FUNNEL CLOUDS"
  data$EVTYPE[grep("FUNNEL", data$EVTYPE)] <- "WALL/FUNNEL CLOUDS"
  data$EVTYPE[grep("DAM", data$EVTYPE)] <- "DAM ACCIDENT"
  data$EVTYPE[grep("SEAS",data$EVTYPE)] <- "ROUGH SEAS"
  data$EVTYPE[grep("WAVE",data$EVTYPE)] <- "ROUGH SEAS"
  data$EVTYPE[grep("WATER",data$EVTYPE)] <- "ROUGH SEAS"
  data$EVTYPE[grep("SWELL",data$EVTYPE)] <- "ROUGH SEAS"
  data$EVTYPE[grep("COOL", data$EVTYPE)] <- "COLD"
  data$EVTYPE[grep("SMOKE",data$EVTYPE)] <- "SMOKE"
  data$EVTYPE[grep("BURST",data$EVTYPE)] <- "DOWNBURST"
  data$EVTYPE[grep("MARINE",data$EVTYPE)] <- "MARINE ACCIDENT"
  data$EVTYPE[grep("HYPOTHERMIA",data$EVTYPE)] <- "HYPOTHERMIA"
  data$EVTYPE[grep("HYPERTHERMIA",data$EVTYPE)] <- "HYPERTHERMIA"
  data$EVTYPE[grep("TURBULENCE",data$EVTYPE)] <- "TURBULENCE"
  data$EVTYPE[grep("SUMMARY", data$EVTYPE)] <- "OTHER"
  data$EVTYPE[grep("TEMPERATURE", data$EVTYPE)] <- "OTHER"
  data$EVTYPE[grep("URBAN", data$EVTYPE)] <- "OTHER"
  data$EVTYPE[grep("NONE", data$EVTYPE)] <- "OTHER"
  data$EVTYPE[grep("RECORD", data$EVTYPE)] <- "OTHER"
  data$EVTYPE[grep("APACHE", data$EVTYPE)] <- "OTHER"
  data$EVTYPE[grep("NO SEVERE", data$EVTYPE)] <- "OTHER"
  data$EVTYPE[grep("PATTERN", data$EVTYPE)] <- "OTHER"
  data$EVTYPE[grep("HIGH", data$EVTYPE)] <- "OTHER"
  data$EVTYPE[grep("HEAVY", data$EVTYPE)] <- "OTHER"
  data$EVTYPE[grep("SOUTHEAST", data$EVTYPE)] <- "OTHER"
  data$EVTYPE[grep("EXCESSIVE", data$EVTYPE)] <- "OTHER"
  data$EVTYPE[grep("SEICHE", data$EVTYPE)] <- "OTHER"
  data$EVTYPE[grep("RED FLAG", data$EVTYPE)] <- "OTHER"
  data$EVTYPE[grep("FLOYD", data$EVTYPE)] <- "OTHER"
  data$EVTYPE[grep("\\?", data$EVTYPE)] <- "OTHER"
```

The financial damage is simply the sum of the property and the crop damages.

```{r}
  data$FDMG <- data$PROPDMG * data$PROPDMGEXP + data$CROPDMG * data$CROPDMGEXP
```

### Computing the mean and the total

The mean, the sum and the number was measured for every type of event.

```{r}
  dataSum <- aggregate(subset(data, select = c("INJURIES", "FATALITIES", "FDMG")), list(data$EVTYPE), sum)
  colnames(dataSum) <- c("EVTYPE", "INJURIESSUM", "FATALITIESSUM", "FDMGSUM")
  dataMean <- aggregate(subset(data, select = c("INJURIES", "FATALITIES", "FDMG")), list(data$EVTYPE), mean)
  colnames(dataMean) <- c("EVTYPE", "INJURIESMEAN", "FATALITIESMEAN", "FDMGMEAN")
  dataFinal <- merge(dataSum, dataMean)
  remove(data, dataSum, dataMean)
```

## Results

We do two different anlysis, first the relation between different variables, and the second we do a rank for each variable.

### Damage-fatalities plot

The damages and fatalies are related as shows the following plot:

```{r}
  library(ggplot2)
  pl <- ggplot(dataFinal, aes(x = FDMGMEAN, y = FATALITIESMEAN)) + geom_point()
  pl <- pl + xlab("damages") + ylab("fatalities")
  print(pl)
```

  There are a lot of financially expensive events that do not cost many lives. And the other way arround.
  In fact this plots show that we can split between financially expensive events and fatal events.

### Damage-injuries plot

The damages and injuries are related as follows:

```{r}
  pl <- ggplot(dataFinal, aes(x = FDMGMEAN, y = INJURIESMEAN)) + geom_point()
  pl <- pl + xlab("damages") + ylab("injuries")
  print(pl)
```

The seperation betweem financial expensive and humanly expensive events is not as clear as before, in fact, there is an event that is both: very financial and humanly expensive.

### Fatalities-injuries plot

The fatalities and injuries are related as follows:

```{r}
  pl <- ggplot(dataFinal, aes(x = FATALITIESMEAN, y = INJURIESMEAN)) + geom_point()
  pl <- pl + xlab("fatalities") + ylab("injuries")
  print(pl)
```

It comes without surprise that there is a relation between fatalities and number of injuries.

### Damage rank

The ranking of financial damage in total:

```{r}
  head(subset(dataFinal[with(dataFinal, order(-FDMGSUM)),],select = c("EVTYPE", "FDMGSUM")), 20)
```

The ranking of financial damage in average:
```{r}
  head(subset(dataFinal[with(dataFinal, order(-FDMGMEAN)),],select = c("EVTYPE", "FDMGMEAN")), 20)
```
As we can see, floods are the events that give more financial damages, however if we compute the mean damage per event, hurricane is by far the worst event. 

### Fatalities rank

The ranking of the number of fatalities in total:

```{r}
  head(subset(dataFinal[with(dataFinal, order(-FATALITIESSUM)),],select = c("EVTYPE", "FATALITIESSUM")), 20)
```

The ranking of the number of fatalities in average:
```{r}
  head(subset(dataFinal[with(dataFinal, order(-FATALITIESMEAN)),],select = c("EVTYPE", "FATALITIESMEAN")), 20)
```

The events with more fatalities are tornados, heat periods and floods.
The mean rank is dominated by very low frequent events and so is not very significant.

### Injuries rank

The ranking of the number of injuries in total:

```{r}
  head(subset(dataFinal[with(dataFinal, order(-INJURIESSUM)),],select = c("EVTYPE", "INJURIESSUM")), 20)
```

The ranking of the number of injuries in average:

```{r}
  head(subset(dataFinal[with(dataFinal, order(-INJURIESMEAN)),],select = c("EVTYPE", "INJURIESMEAN")), 20)
```

The events with more injuries are tornados, storms, heat periods and floods.
The mean rank is dominated by very low frequent events and so is not very significant.
